---
title: "Housing affordability index for professors"
format: html
editor: visual
---

For this assignment, I continue my investigation of professor salaries. In particular, I investigate the relative 'impoverishment' of faculty as median home prices rise from 2018-2022, disproportionately to salary increases.

Data on Q3 median home values for 2018-2022 was extracted from:

[https://www.zillow.com/research/data/]{.underline}

Data on 2022 salaries was extracted from:

[https://www.ziprecruiter.com/Salaries/What-Is-the-Average-Professor-Salary-by-State]{.underline}

For 2018-2021 salaries, average professor pay raise was extracted from:

[https://www.aaup.org/sites/default/files/AAUP-2021-SurveyTables.pdf]{.underline}

From the table, raises were:

2022: 2%

2021: 1%

2020: 2.8%

2019: 2%

2018: 2.8%

These were used to reduce 2022 pay to calculate 2018-2021 pay.

### Data manipulation

\>For each year a basic affordability index defined as home price to annual income ratio was calculated. This was inspired by

[https://data.oecd.org/price/housing-prices.htm]{.underline}

Basically, a home price of 240K relative to a salary of 80K would give an index of 3.

\>A simple difference (change between 2022 and 2018) was also calculated. for example, a difference of 1 would indicate a home in 2022 would cost an extra yearly salary, as compared to 2018.

\>A percentage increase in monthly expenditure on housing was also calculated. This is the percent change between 2018 and 2022 for mortgage/ monthly salary. Mortgage was a calculated as 2\*price/360, monthly salary as salary/12.

### Table

The table show the following:

\>State, Median Salary, Home Prices for 2018 and 2022, the Affordability Index for 2022, and the Change in Affordability between 2018-2022.

\>The long table is put in a scroll box

\>The table is themed

\>The headers are nested

\>The rows are highlight-able

\>The Change in Affordability is color-coded for a difference greater than 1 salary

\>A tooltip over Change in Affordability shows the percentage increase in monthly expenditure 2018-2022

\>A sparkline 'Trend' shows yearly Affordability Indices for 2018-2022 when hovered over

Clearly, meager yearly raises - often the case in academia as raises are tied to promotions and less so performance - mean that the real earnings of university professors have decreased dramatically as housing costs increase!

```{r}
#| warning: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(sparkline)
library(tidyr)
library(dplyr)
library(htmlwidgets)
```

-   

```{r}
table <- read.csv("statedatafortable.csv")
colnames(table)[1] <- gsub('^...','',colnames(table)[1])
```

-   

```{r}
tidytable <- read.csv("statedataforsparkle.csv") %>%
    gather(key ="Quarter", value ="Index", Q3_2018:Q3_2022)
colnames(tidytable)[1] <- gsub('^...','',colnames(tidytable)[1])
```

-   

```{r, results='hide'}
table.kable <- as.data.frame(table)
table.kable %>% add_column(Trend = NA)
```

-   

```{r}
for(s in table.kable$State){
  # vector of data for figure for cases
  trend <- tidytable %>%
    filter(State==s) %>%
    arrange(Quarter) %>%
    pull(Index)
  # add figure to data.frame
  table.kable[which(table.kable$State==s),"Trend"] <- spk_chr(trend)
}
```

-   

```{r}
table.kable$Difference.22.18 <- ifelse(
  table.kable$Difference.22.18 > 1,
  cell_spec(table.kable$Difference.22.18, color = "red", bold = T, 
            tooltip=paste("Professors in", table$State, "pay",table$Percent, "% of their income more on housing in 2022")),
  cell_spec(table.kable$Difference.22.18, 
            tooltip=paste("Professors in", table$State, "pay",table$Percent, "% of their income more on housing in 2022"))
)
```

-   

```{r}
sparkline(0)
table.kable.full <- table.kable%>% 
  select(State, Salary.2022, Home.Price.2018, Home.Price.2022, Index.2022, Difference.22.18, Trend) %>%
  kbl(escape = FALSE,
      col.names = c("State",
                    "Median Salary",
                    "2018",
                    "2022",
                    "2022",
                    "Change, 2018-22", 
                    "Trend"),
  align = c("l","r","r","r","r","r","r"),
      digits = 1, format.args = list(big.mark = ",")) %>%
  add_header_above(c(" ", " ", "Median Home Price" = 2, "Affordability Index" = 2, " "))  %>%
  kable_styling(font_size=15, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "400px") %>%
  kable_paper(lightable_options = "hover", full_width = FALSE)
table.kable.full
```
